`r if(knitr:::is_latex_output()) '\\startappendices'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

<!-- If you feel it necessary to include an appendix, it goes here. The first appendix should include the commands above. -->


# Appendix - Code

This first appendix includes the non-stationary gaussian processes in Gpytorch:

**In 02-rmd-basics-code.Rmd**

```{r ref.label='chunk-parts', eval=FALSE, echo = TRUE}
```

**And here's another one from the same chapter, i.e. Chapter \@ref(code):**

```{r ref.label='oxford-logo-rotated', eval=FALSE, echo = TRUE}
```



# Appendix - Figures

\setcounter{figure}{0}
\makeatletter 
\renewcommand{\thefigure}{S\@arabic\c@figure}
\makeatother

<!-- FIGURE S2 -->
\begin{figure}[!h]
	\centering
	\includegraphics[width=1\linewidth]{figures/extra/figureS2.png}
	\caption{XXXX}
	\label{fig:figureS02}
\end{figure}


<!-- FIGURE S3 -->
\begin{figure}[!h]
	\centering
	\includegraphics[width=1\linewidth]{figures/extra/figureS3.png}
	\caption{XXXX}
	\label{fig:figureS03}
\end{figure}


